Unshuffling Permutations

### 2015-07-01 article.tex
Partie algebre (suite)

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, or $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{proposition}
\begin{proof}
    To prove that, for $j = 1, 2, 3$, $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    \smallskip

    Let $\pi$ be a permutation. If $P$ is a set of indexes of letters of
    $\pi$, we denote by $\widetilde{\pi}$ the set
    $\{|\pi| - i  + 1 : i \in P_1\}$. Now since this operation
    $\widetilde{\,}$ define a bijection on the set of the subsets of
    $[|\pi|]$ and the standardisation operation commutes with the
    mirror operation on words, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]} \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

    ADAPT NEXT ARGUMENTS AND FINISH.

    First of all, since $\pi$ is a square permutation and $\rho$ is a
    square root of $\pi$, $\rho \otimes \rho$ has a nonzero coefficient
    in the expansion of $\Delta(\pi)$. This implies that there are two
    disjoint sets of indexes $P_1$ and $P_2$ such that
    $P_1 \uplus P_2 = [|\pi|]$ and, by denoting $\pi_{|P_1}$ by $u$ and
    $\pi_{|P_2}$ by $v$,
    \begin{equation} \label{equ:square_stability_same_std}
        \STD(u) = \rho = \STD(v).
    \end{equation}
    \smallskip

    To prove~\ref{item:square_stability_1}, set
    $\widetilde{P_1} := \{|\pi| - i  + 1 : i \in P_1\}$,
    $\widetilde{P_2} := \{|\pi| - i  + 1 : i \in P_2\}$,
    $u' := \widetilde{\pi}_{|\widetilde{P_1}}$, and
    $v' := \widetilde{\pi}_{|\widetilde{P_2}}$.
    Now, by construction, $\widetilde{u'} = u$ and $\widetilde{v'} = v$.
    Then, by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \widetilde{\rho} = \STD(v')$, so that $\widetilde{\rho}$
    is a square root of $\pi$.
    \smallskip

    Let us establish~\ref{item:square_stability_2}. Set
    $u' := \bar \pi_{|P_1}$ and $v' := \bar \pi_{|P_2}$. By definition
    of the operation $\bar\,$, for all indexes $i$ and $j$, we have
    $u_i < u_j$ if and only if $\bar u'_i > \bar u'_j$, the same
    property holding for $v$ and $v'$. Then,
    by~\eqref{equ:square_stability_same_std},
    $\STD(u') = \bar \rho = \STD(v')$, showing that $\bar \rho$ is a
    square root of $\bar \pi$.
    \smallskip

    Finally, let us show~\ref{item:square_stability_3}. Set
    $P^{-1}_1 := \{u_i : i \in [|u|]\}$,
    $P^{-1}_2 := \{v_i : i \in [|v|]\}$, $u' := \pi^{-1}_{|P^{-1}_1}$,
    and $v' := \pi^{-1}_{|P^{-1}_2}$.

    TO CONTINUE
\end{proof}
\medskip

>>>>

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, and $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{proposition}
\begin{proof}
    To prove that, for $j = 1, 2, 3$, $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    In the sequel, $\pi$ is a permutation.
    \smallskip

    If $P$ is a set of indexes of letters of $\pi$, we denote by
    $\widetilde{\pi}$ the set $\{|\pi| - i  + 1 : i \in P\}$. Now since
    this operation $\widetilde{\,}$ define a bijection on the set of the
    subsets of $[|\pi|]$ and the standardisation operation commutes with
    the mirror operation on words, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \widetilde{\STD\left(\pi_{|P_1}\right)}
        \otimes \widetilde{\STD\left(\pi_{|P_2}\right)} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

    Next, since by definition of the complementation operation on
    permutations, for any permutation $\tau$ and any indexes $i$ and $k$,
    we have $\tau_i < \tau_j$ if and only if $\bar \tau_i > \bar \tau_j$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_2(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_2(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_2(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\bar \pi_{|P_1}\right)
        \otimes \STD\left(\bar \pi_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_2\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_2\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_2 \otimes \phi_2) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_2$ is a morphism of coalgebras and hence, that
    $\phi_2$ is a morphism of associative algebras.
    \smallskip

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau_i : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection and
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
\end{proof}
\medskip
